From big data to deep insight in developmental science

### 2015-08-03 wiley-big-data-devel/big-data.tex
nearly complete draft

There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.

There are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) \cite{MGRS} collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
Launched in 2011, the Japan Environment and Children's Study \cite{JECS} is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development. 
Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
Health outcomes and exposure measurements will continue until the participating children become 13 years old.
The study is in its early stages.
The study has produced a handful of publications \cite{JECS_pubs}, and no data sharing plans have been announced.
Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health & Development \cite{NSHD_UK} and the National Child Development Study (NCDS) \cite{NCDS_UK}.
NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
Measures include genetic, health, education, social, and economic factors.
The study is funded by the British government and data are available by registration from the U.K. Data Service \cite{UK_cohort_data}.
The British Cohort Studies have generated more than 3,500 publications to-date \cite{UK_cohort_citations}.
Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset \cite{PISA}, which generates more than 5,000 search hits.

@@

The number and variety of topics studied

The Panel Study on Income Dynamics (PSID) \cite{PSID} is another large-scale population based effort.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
The study consists of survey responses to questions about employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, education, and other topics. As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}. 
Much of data is available publicly to users who register (http://simba.isr.umich.edu/U/ca.aspx) subject to the provisions of a formal set of use conditions (http://simba.isr.umich.edu/U/CondUse.aspx) that focus largely on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data. 
In addition, some restricted data elements are only available subject to a formal restricted use contract (http://simba.isr.umich.edu/restricted/RestrictedUse.aspx) between the investigator and the data owner, the University of Michigan.

The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} was a survey designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year. 
The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
The Ad Health cohort has been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32. 
In addition to survey data Add Health includes biomarker data from a blood test taken during Wave IV. 
Data are available about glucose, hemoglobin (HbA1c) levels, and blood lipids. 
The Add Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.

In comparison, the NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences, and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
Unlike the Ad Health study, data in SECCYD are not available to the public, but are restricted.
Access is granted only by application, and approval is granted on a case-by-case basis.
The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.

The purpose of this project is to study the early development of a group of children who are at risk regarding later successful adjustment and for whom we have little information: children living in rural, largely poor communities. 
A birth cohort of 800 children in three rural counties of North Carolina and 600 children in three rural counties of Pennsylvania were studied over the first 3 years of life.
https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=&amp;q=Bayley&amp;searchSource=revise

@@

Large data samples have long been required in behavior genetics order to detect what can often be small effect sizes.
Several large-scale genetics studies that are not specifically developmental deserve mention.
The Genome of the Netherlands Project(http://www.bbmri.nl/en-gb/activities/rainbow-projects/131-genome-of-the-netherlands) is a publicly available dataset from n=250 families, consisting of two parents plus one of their adult children \cite{Genome_Netherlands}.
The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia (http://gemstudy.org/main/index.html).
The Psychiatric Genomics Consortium (PGC) (http://www.med.unc.edu/pgc) contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (http://www.broadinstitute.org/mpg/ricopili/).

The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} (https://en.wikipedia.org/wiki/Twins_Early_Development_Study) contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
DNA samples from more than 5,000 twins pairs are also available.
TEDS has generated more than 350 publications to-date.
Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables that needed satisfy the aims. 
Proposals that do not overlap with analyses already being planned or carried out by the TEDS research team or other collaborators are usually approved.

The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} focuses on genetic and environmental ontributions to measures of family relationships and mental health.
TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (n=3,000 individuals) and their families.
While still relatively early in data collection, more than 24 publications reference the TOSS dataset.
Data sharing plans have not yet been announced.

Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
Researchers focused on survey, self-report, and experimenter ratings (from video) of family relationships, adolescent adjustment and competence.
A Google Scholar search shows 73 publications to-date.
The data will be shared by request to the project team.
The team plans to share the data openly later in 2015.

Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
Since its inception in 1976, the Colorado Adoption Project (CAP) (http://ibg.colorado.edu/cap/index.html) has enrolled more than 2,400 participants from more than 450 families.
Data include cognitive abilities, temperament measures, demographic variables, 
Additionally parents completed a questionnaire assessment which included measures of temperament
CAP has generated more than 200 published articles (http://ibg.colorado.edu/cap/publishedarticles.html).
Data collection continues.
The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
EGADS is a prospective adoption study of (n=561) triads of birth parents, adoptive parents, and adopted children in the U.S. 
The study focuses on genetic and non-genetic factors that influence children’s adjustment from infancy to age 9. 
The data include self- and other-report measures, DNA, and salivary cortisol.
Data are available by request, with plans for more open data sharing in the near future.

@@

Several large-scale studies have focused on developmental patterns in brain structure or activity.
For practical reasons -- motion artifacts make imaging very young children extremely challenging -- most projects have focused on children and adolescents.
One of the earliest studies of its kind was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
More than 100 of the participants were scanned on more than one occasion, with the time between scans approximately two years.
The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data appear not to be available for analyses by researchers outside the original investigative team.

@@

\subsection{Behavior}

In the realm of behavior, there are a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.
The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
The Bayley is widely used in clinical and epidemiological settings such as the EPICure <http://www.epicure.ac.uk/> because of the norms.
However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE), 1972-1992 project on ICPSR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/4091?sortBy=19&amp;q=Bayley&amp;searchSource=revise).
The data are freely available to registered users of ICSPR.

The Wechsler intelligence tests were developed by Dr. David Wechsler, beginning in the 1930's. The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
The current version (WISC V) is provided by Pearson (http://www.pearsonclinical.com/psychology/products/100000771/wechsler-intelligence-scale-for-childrensupsupfifth-edition--wisc-v.html).
Data about the norms themselves appears not to be available to researchers.
However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR (https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/13604?q=wisc&amp;searchSource=find-analyze-home&amp;sortBy=). 
The data are restricted, and thus are only available by specific authorization.

The Family Life Project also collected and reported WPPSI data.

\subsubsection{Language}

There are several large-scale datasets associated with language processing.
The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established.
It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, some of them publicly.

The Human Speechome Project (http://www.media.mit.edu/cogmac/projects/hsp.html) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
The project has generated more than 78 publications (http://www.media.mit.edu/cogmac/publications.html), but the data have not been made available outside of the original investigative team.
Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.

% edit previous
WordBank (http://wordbank.stanford.edu/) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from a longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.

The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed (http://www.lenafoundation.org/customer-resources/videos).
LENA stands for Language ENvironment Analysis.
The system provides an automatic language collection and analysis tool for speech language professionals and parents.
The system was designed and is distributed by the LENA Foundation.

\subsubsection{Vulnerable Populations}

%- Infant Behavior Questionnaire, CBC
While no archive for IBQ data across studies exists, IBQ data on 1,388 participants (n = 658) who were group matched within site with a group of non-exposed comparison infants (n=730) are archived at ICSPR (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34312?sortBy=19&amp;q=infant+behavior+questionnaire&amp;searchSource=revise) as part of the Maternal Lifestyle Study in Four Sites in the United States, 1993-2011.

MSL was a large NIH longitudinal studies of children with prenatal cocaine exposure (PCE). MLS was a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
IBQ data are also available from ICPSR from the Family Life Project (http://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=19&amp;q=ibq&amp;searchSource=revise).

%  * [Family Life Project](http://flp.fpg.unc.edu/)

The purpose of this project is to study the early development of a group of children who are at risk regarding later successful adjustment and for whom we have little information: children living in rural, largely poor communities. 
A birth cohort of 800 children in three rural counties of North Carolina and 600 children in three rural counties of Pennsylvania were studied over the first 3 years of life.
https://www.icpsr.umich.edu/icpsrweb/ICPSR/studies/34602?sortBy=&amp;q=Bayley&amp;searchSource=revise

MacArthur Network on Transitions to Adulthood (http://www.transitions2adulthood.com/). Data not obviously open or available.
Research Network on Adolescent Development & Juvenile Justice - See more at: https://www.macfound.org/networks/research-network-on-adolescent-development-juvenil/#sthash.PQOwqv14.dpuf
Research Network on Youth and Participatory Politics - See more at: https://www.macfound.org/networks/research-network-on-youth-and-participatory-politi/#sthash.egU1Wwvy.dpuf

* Bucharest Early Intervention Project (BEIP), <http://www.childrenshospital.org/beip>
* NPR story, <http://www.npr.org/blogs/health/2014/02/20/280237833/orphans-lonely-beginnings-reveal-how-parents-shape-a-childs-brain>

@@

%LearnLab, Edith Gumm's new gig
The Trends in International Math and Science Study (TIMSS) studied mathematics and science achievement in 4th and 8th grade for more than 20 years.
Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
Both studies provide data available for public download (http://timss.bc.edu/).
The 1999 TIMSS video study examined teaching practices in seven countries, and some of the materials are available for public use (http://www.timssvideo.com/).
The Measures of Effective Teaching (MET, http://www.metproject.org/) Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
The MET video data are stored and shared at ICSPR.

\subsection{Commercially Initiated and Managed Datasets}

The College Board shares SAT and college cost and scholarship data with the research community by application (http://www.collegeboard.com/inquiry/csinquiry_form.html).
The ACT (http://www.act.org/research/). Privacy policies (http://www.act.org/privacy.html).

Google, Twitter, Amazon, Facebook, SnapChat.

@@

\subsubsection{File and Data Formats}

>>>>

There are several large government-collected and shared datasets with information about employment patterns and educational attainment in children and youth.
The Bureau of Labor Statistics reports data about youth employment and unemployment patterns derived from the Census Bureau's Current Population Survey (http://www.bls.gov/cps/demographics.htm).
The U.S. Department of Education's National Center for Educational Statistics (NCES) collects and analyzes statistics on American primary, secondary, and post-secondary education, and many of the datasets are either public; data with individually identifiable data are available to researchers under a restricted use agreement.
The Trends in International Math and Science Study (TIMSS) has studied mathematics and science achievement in 4th and 8th grade for more than 20 years \cite{TIMSS}.
TIMSS is funded by the U.S. Department of Education, with domestic and international operations coordinated by a consortium of academic, association, and commerical partners.
The 2001 TIMSS assessment involved more than 20,000 U.S. students from more than 1,000 schools plus data from more than 60 countries.
The next TIMSS data collection is in 2015 and will include students in grades 4, 8 and 12.
Its sister study, the Progress in International Reading Literacy Study (PIRLS) has been studying reading achievement at the 4th grade.
Both studies provide data available for public download.

In addition to TIMSS, there are several examples of large-scale longitudinal studies focused on children's growth, health, and education from outside the U.S.
The World Health Organization's (WHO) Multicentre Growth Reference Study (MGRS) \cite{MGRS} collected growth and related data from 8,500 children from widely different ethnic backgrounds and cultural settings, specifically Brazil, Ghana, India, Norway, Oman, and the U.S.
MGRS data provide body mass index, height for children 5-19, and weight data norms for children 5-10, in addition to other anthropomorphic measurements.
Launched in 2011, the Japan Environment and Children's Study \cite{JECS} is a birth cohort study involving 100,000 parent-child pairs with the goal of evaluating the impact of various environmental factors on children's health and development. 
Pilot studies started in 2008 and the hundred thousandth participating mother was enrolled in March 2014. 
Health outcomes and exposure measurements will continue until the participating children become 13 years old.
The study is in its early stages.
The study has produced a handful of publications \cite{JECS_pubs}, and no data sharing plans have been announced.
Several large-scale birth cohort studies have been undertaken in the U.K., including the National Survey of Health & Development \cite{NSHD_UK} and the National Child Development Study (NCDS) \cite{NCDS_UK}.
NCDS is an ongoing longitudinal study that seeks to follow the lives of all those living in Great Britain who were born in one particular week in 1958.
Originally focused on improving infant mortality, the study has expanded to focus on factors predicting wellbeing across the lifespan.
Measures include genetic, health, education, social, and economic factors.
The study is funded by the British government and data are available by registration from the U.K. Data Service \cite{UK_cohort_data}.
The British Cohort Studies have generated more than 3,500 publications to-date \cite{UK_cohort_citations}.
Finally, the Organization for Economic Cooperation and Development (OECD) publishes the PISA comparative educational dataset \cite{PISA}, which generates more than 5,000 search hits.

@@

Investigator-managed big data projects cover a wide range of topics and and even wider range of research methodologies.
In several cases, the studies involve partnerships between government agencies and academic investigators.
The Panel Study on Income Dynamics (PSID) \cite{PSID} is large-scale population based survey study focusing on employment, income, wealth, expenditures, health, marriage, childbearing, child development, philanthropy, and education.
PSID began in 1968 with a U.S. nationally representative sample of 5,000 families and more than 18,000 individuals.
As of mid-2015, PSID had generated more than 3,900 publications \cite{PSID_pubs}. 
Much of data is available publicly to registered users and agree to the terms of use conditions \cite{PSID_term} that focus on ensuring that participants are not reidentified and that users properly cite and acknowledge use of the data. 
In addition, some restricted data elements -- e.g., geospatial data, school-level identifiers -- are only available subject to a formal restricted use agreement \cite{PSID_restricted_use} between the investigator and the data owner, the University of Michigan.

The National Longitudinal Study of Adolescent Health (Ad Health) \cite{adolescent_health} began as a survey study designed to measure the multiplicity of influences on adolescents' health and unhealthy behaviors. 
Undertaken in response to a Congressional mandate, the study was carried out in two phases, the first involving surveys of 90,000 7-12th graders at 145 schools around the U.S. during the 1994-1995 school year. 
The second phase involved 20,000 participants, where both the target adolescents and their parents underwent interviews at two time points separated by a year. 
The Ad Health cohort has subsequently been followed into young adulthood with four in-home interviews, the most recent in 2008, when the sample was aged 24-32.
During Wave IV, biomarker data from a blood test has made available data about glucose, hemoglobin (HbA1c) levels, blood lipids, and related measures. 
The Ad Health data \cite{harris_add_health} are archived at ICSPR \cite{ICPSR} and are available for public use.
As of mid-2015, more than 5,000 publications \cite{add_health_publications} had arisen from the Add Health dataset.

The NICHD Study of Early Child Care and Youth Development (SECCYD) prospectively followed the experiences of 1,364 infants born at 10 locations throughout the United States in 1991.
The goal was to learn more about the kind and quality of child care experiences and the effects of different child care settings on developmental outcomes. 
Measures included video, surveys and interviews, behavioral tests, biomarkers such as salivary cortisol, blood pressure, and anthropometric measures, and demographic information.
The study included multiple respondents, including the child, parents, caregivers, clinicians, and child care center directors.
Unlike the Ad Health study, data in SECCYD access are restricted, available only by application, and approval is granted on a case-by-case basis.
Staff at the National Institute of Child Health and Human Development manage the data and access approval process.
The SECCYD has generated more than 214 publications as of mid-2015 \cite{seccyd_publications}.

The Family Life Project (FLP) studied the early development children living in rural, largely poor communities in North Carolina and Pennsylvania. 
A birth cohort of 800 children were studied over the first 3 years of life in a series of home, child care visits and phone calls when the infants were 2, 6, 15, 24, and 36 months of age. 
Data collected included interviews, questionnaires, videotaped interactions, and biological samples yielding information about economic and community factors, family, home, language and cognition, stress, and genotype.
These features make the FLP large in volume, velocity, and variety.
Data from Phase I of the FLP are archived at ICPSR \cite{Family_Life_data}, but are subject to restrictions on access to the sensitive and identifiable nature of the data.
The FPL site lists about 50 publications through 2013.

@@

Large data samples have long been required in behavior genetics order to detect what can often be small effect sizes.
Several large-scale genetics studies that are not specifically developmental deserve mention.
The Genome of the Netherlands Project(http://www.bbmri.nl/en-gb/activities/rainbow-projects/131-genome-of-the-netherlands) is a publicly available dataset from n=250 families, consisting of two parents plus one of their adult children \cite{Genome_Netherlands}.
The Genes, Environment, and Melanoma Study (GEMS) focuses on risk for melanoma in a sample of 3,700 individuals from the North America, Europe, and Australia \cite{GEMS}.
The Psychiatric Genomics Consortium (PGC) \cite{PGC} contains data from more than 170,000 individuals who have psychiatric diagnoses or are considered at-risk.
Many disorders featured (autism, attention-deficit hyperactivity disorder, and schizophrenia) in the PGC dataset have developmental dimensions.
Genome-wide association study (GWAS) results from the PGC dataset may be viewed in a specialized web browser tool (http://www.broadinstitute.org/mpg/ricopili/).

The U.K.-based Twins Early Development Study (TEDS) \cite{TEDS} contains survey data from 15,000 English and Welsh families who gave birth to twins between 1994 and 1996.
Some 13,000 pairs remain actively involved, providing lab-based behavioral task and survey responses.
DNA samples from more than 5,000 twins pairs are also available.
TEDS has generated more than 350 publications to-date.
Researchers wishing to gain access to the data may do so by submitting a formal data request specifying the aims of the planned research and the variables that needed satisfy the aims. 
Proposals that do not overlap with analyses already being planned or carried out by the TEDS research team or other collaborators are usually approved.

The Twin and Offspring Study in Sweden (TOSS) \cite{TOSS} is another twin-focused study aimed at uncovering genetic and environmental ontributions to measures of family relationships and mental health.
TOSS extends the Twin Moms Project to include a sample of twin fathers (N=320 twin pairs and their families).
In addition, TOSS plans to collect another 250 pairs of twin mothers for a target sample of 900 twin parents (n=3,000 individuals) and their families.
While still relatively early in data collection, more than 24 publications reference the TOSS dataset.
Data sharing plans have not yet been announced.

Adoption studies provide a unique window on the mix of genetic and non-genetic factors that influence child development.
Illustrative U.S. examples include the Colorado Adoption Project (CAP), the Nonshared Environment in Adolescent Development (NEAD) project, and the Early Growth and Development Study (EGADS).
Since its inception in 1976, the Colorado Adoption Project (CAP) \cite{CAP} has enrolled more than 2,400 participants from more than 450 families.
Data include cognitive abilities, temperament measures, and demographic variables.
CAP has generated more than 200 published articles \cite{CAP_pubs}.
Data collection continues.
The study aims to be the definitive genetically informed longitudinal study of lifespan development \cite{rhea_colorado_2013}.
Data from 1976-1989 are archived at Dataverse (https://dataverse.harvard.edu/dataset.xhtml?persistentId=hdl:1902.1/00913) and can be accessed by application to Dataverse.
NEAD studied a sample of 720 families, including twins, siblings, and their parents, in a longitudinal design that involved up to 3 data collection waves.
Researchers focused on survey, self-report, and experimenter ratings (from video) of family relationships, adolescent adjustment and competence.
A Google Scholar search shows 73 publications to-date.
The data will be shared by request to the project team.
The team plans to share the data openly later in 2015.
EGADS is a prospective adoption study of (n=561) triads of birth parents, adoptive parents, and adopted children in the U.S. 
The study focuses on genetic and non-genetic factors that influence children’s adjustment from infancy to age 9. 
The data include self- and other-report measures, DNA, and salivary cortisol.
Data are available by request, with plans for more open data sharing in the near future.

@@

Several large-scale studies have focused on developmental patterns in brain structure or activity.
One of the earliest studies of its kind was a combined longitudinal and cross-sectional study of child and adolescent structural brain development led by Jay Giedd at NIMH \cite{giedd_brain_1999}.
The study collected structural brain information from 145 4 to 20-year-olds along with behavioral and clinical data.
More than 100 of the participants were scanned on more than one occasion, with the time between scans approximately two years.
The original paper describing this study has been cited more than 3,300 times, as of mid-2015.
The dataset has generated other highly cited papers \cite{gogtay_dynamic_2004}, but the data appear not to be available for analyses by researchers outside the original investigative team.

@@

\subsubsection{Language, Cognition, Temperament}

The child language community has pioneered the aggregation and sharing of datasets.
The CHILDES/TalkBank \cite{macwhinney_childes_2001} archive is one of the largest and most well-established.
It consists of transcripts and audio and video recordings of children's utterances along with recent data from adults with aphasia.
CHILDES/TalkBank has generated more than ten thousand of citations, according to Google Scholar, and the data are available to the research community, many of them publicly.

Other large-scale sources of language-related data are more recent.
WordBank (http://wordbank.stanford.edu/) is an open access archive that consists of data from more than 40,000 adminstrations of the MacArthur-Bates Communicative Development Inventory (CDI).
Catherine Tamis-Lemonda has released video data (https://nyu.databrary.org/volume/8) from an NSF-funded longitudinal study of child language consisting of more than 1,000 sessions of infants and children (9 mos-7.6 years) and their mothers carrying out a series of semi-structured tasks.
The Human Speechome Project (http://www.media.mit.edu/cogmac/projects/hsp.html) at MIT's Media Lab involved recording 10 hours of audio and video from one child's home on a daily basis from birth to age three.
The project has generated more than 78 publications (http://www.media.mit.edu/cogmac/publications.html), but the data have not been made available outside of the original investigative team.
Data was gathered at an average rate of 200 gigabytes per day, necessitating the development of sophisticated data-mining tools to reduce analysis efforts to a manageable level, and transcribing significant speech added a labor-intensive dimension.
The LENA Foundation has created a technology framework that allows children's speech in natural settings to be recorded and analyzed \cite{LENA}.
LENA stands for Language ENvironment Analysis.
The system provides an automatic language collection and analysis tool for speech language professionals and parents.
The system was designed and is distributed by the LENA Foundation.
No large-scale archive for LENA data currently exists, but some researchers who use the tool have begun to explore the creation of a data repository specialized for these sorts of data.

In other areas of cognition, a number of measures that have become standards whose widespread adoption has resulted in data that are large in volume, velocity or variety.
Davida Teller pioneered empirical techniques for measuring visual acuity in preverbal children, and the use of Teller Acuity Cards has resulted in the publication of age-based norms for visual development in infants to four-year-olds \cite{mayer_monocular_1995} based on a sample of more than 400.
A companion study on children in Brazil \cite{salomao_large_1995} was conducted with an even larger sample.

The Bayley Scales of Infant and Toddler Development \cite{bayley2006bayley} arose from the Nancy Bayley's Berkeley Growth Study \cite{jones_berkeley_1941} that studied the behavior, motor coordination, physical activity, cognitive function, and maternal behavior of 61 healthy newborn infants.
The California First Year Mental Scale \cite{bayley1933california} and The California Infant Scale of Motor Development \cite{bayley1936california} the precursors.
The Bayley is widely used in clinical and epidemiological settings such as the EPICure \cite{EPICure} because of the norms even though there are questions about its validity \cite{hack_poor_2005}.
However, data about those norms appears to be under the control of the Pearson Publishing company that publishes the Bayley and licenses its use.
Bayley score data are available from the Carolina Abecedarian Project and the Carolina Approach to Responsive Education (CARE) project on ICPSR \cite{CARE}.
The data are freely available to registered users of ICSPR.

The Wechsler intelligence tests were developed by Dr. David Wechsler, beginning in the 1930's. 
The Wechsler Intelligence Scale for Children-III (WISC-III) is designed for children ages 6 - 16, while the Wechsler Preschool and Primary Scale of Intelligence-R (WPPSI-R) is designed for children age 4 - 6 1/2 years.
The current version (WISC V) is provided by Pearson \cite{WISC_V}.
Data about the norms themselves appears not to be available to researchers.
However, data from the Project on Human Development in Chicago Neighborhoods (PHDCN) that collected WISC-R data on 6,000 children, adolescents, and young adults is shared at ICSPR \cite{PHDCN}. 
The data are restricted, and thus are only available by specific authorization.

In the realm of emotion and temperament, two academic-investigator-initiated measures have been widely adopted.
The Infant Behavior Questionnaire (IBQ), a parent survey measure, and the Laboratory Temperament Assessment Battery (Lab-TAB) \cite{Lab_TAB} an observational battery, using a set of 3-5 min episodes mimicking situations in everyday life.
While no archive for IBQ or Lab-TAB data across studies exists, IBQ data on 1,388 participants are archived at ICSPR as part of the Maternal Lifestyle Study (MLS) \cite{MSL}, a longitudinal multi-site observational study of the long-term effects of in-utero exposure to cocaine on child development.
IBQ and Lab-TAB data are also available from ICPSR from the Family Life Project.

@@

There are several investigator initiatiated or managed education studies that deserve mention because of their volume, velocity, or variety or because of their availability for reuse by others.
The 1999 TIMSS video study recorded teaching practices in seven countries, and some of the materials are available for public use \cite{TIMSS_video}.
The TIMSS video study repository is hosted at UCLA with support from the Carnegie Foundation for the Advancement of Teaching, The William and Flora Hewlett Foundation and the Gabriella and Paul Rosenbaum Foundation.
The Gates Foundation-funded Measures of Effective Teaching (MET) \cite{METProject} Project videotaped more than 3,000 classroom teachers beginning in 2009, with follow-on studies continuing today.
The MET video data are stored and shared at ICSPR.
The Spatial Intelligence and Learning Center (SILC) \cite{SILC} is an NSF-funded initiative designed to ``develop the science of spatial learning and to use this knowledge to transform educational practice, supporting children and adults in acquiring the scientific, technical, engineering, and mathematical (STEM) skills required for effective participation in an increasingly high-technology society and global economy.''
The SILC website hosts links to a set of test and instruments useful in research on spatial learning.
LearnLab \cite{LearnLab}, another NSF-funded center, combines cognitive theory and computational modeling to understand changes in student knowledge in the context of computer-based math, science and language courses. 
LearnLab hosts a repository, DataShop \cite{DataShop}, to store and openly share detailed data from classrooms employing cognitive tutors.
DataShop enables data storage, management, visualization and analysis in a web-based tool.

Reviewing the set of investigator initiated or managed datasets, we see examples of datasets large in volume, velocity, and variety.
Measures employed are also diverse, from surveys, to observational measures, including video recordings, to both free academic and commercial population-normed test instruments, to biological measurements of physiology, brain structure or function.
The extent to which data are available for secondary reuse, and the process for acquiring access is more variable than for datasets initiated and managed by government entities.

\subsection{Non-Academically Initiated or Managed Sources of Big Data}

Finally, we mention briefly some large-volume sources of developmental data that are collected and managed by private, non-academic or government entities.
More than 1.6 million high school students \cite{lewin_more_2013} take standardized tests or provide financial aid information via measures developed and managed by The College Board or the ACT, both of which are private entities.
The College Board shares Scholastic Aptitude Test (SAT) and college cost and scholarship data with the research community by application \cite{College_Board_data}.
So does the ACT \cite{ACT_research}.

Internet-based service providers operate at an even larger scale.
Google's Gmail has more than 900 million users worldwide \cite{lardinois_gmail_2015}, while Facebook has more than a billion \cite{facebook_2014}.
According to the YouGov site \cite{YouGov}, some 17\% of Gmail users in the United Kingdom are 17-24 years of age.
Facebook's policies require that users be at least 13 years of age \cite{facebook_under_13}, but detailed information about user demographics for Facebook or other social media popular among children and adolescents is not openly available \cite{BI_social_media}.
Of course, detailed information about users, their characteristics, and preferences is the primary asset social media companies mine and market to the business community.
Users receive free services in exchange for providing these data.
Both Google \cite{Google_Research} and Facebook \cite{Facebook_Research} have arms that both conduct research and cooperate with academic researchers, but there has been significant public criticism about the ethics of certain research projects \cite{meyer_everything_2014}.
The primary criticism concerns whether Facebook users had given informed consent to participate in the manipulation of their newsfeeds in the way that researchers whose work is supervised by an IRB would be required to secure if a similar study were taken in a laboratory context.

The scale of data collected and managed by non-academic entities dwarfs that of government or academically-managed initiatives.
Because the data are collected for proprietary business purposes, it is difficult to assess their current or potential impact on the study of human development, but in the final section we turn to a number of technical, conceptual, and theoretical issues that bear on the question.

@@

\subsubsection{Collection from Multiple Sources and in Diverse Formats}

Developmental scientists collect data from sources representing multiple levels of analysis.
Increasingly, the devices used to make measurements provide data and metadata in a structured, organized, machine-readable formats.
